---
title: "Precipitation Affinity for Puerto Rican Trees"
author: "Bob Muscarella"
date: "5/16/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
The idea of this script is to take point occurrence records of tree species, extract values of precipitation at those points, and summarize those data to represent some kind of 'precipitation affinity' for different species.  We start with looking at occurrences in PR and comparing with PR climate data.  Then we get global occurrences from GBIF and BIEN and compare these occurrences with global climate data from [Chelsa](http://chelsa-climate.org/).


## Getting started
First load the packages we need to use and set a working directory.  All of these can be installed as normal via CRAN except for one, `chelsaDL`, which you need to install from GitHub using the `devtools` package like this: 

`devtools::install_github("matthewkling/chelsaDL")`

```{r start, message=F, eval=F}
library(raster)
library(rgbif)
library(CoordinateCleaner)
library(chelsaDL)
library(stringr)
library(BIEN)

# Set an empty working directory
setwd("~/Downloads")
```

## Download all the data
I put a zip file on Github with some of the data used below. You download that with the next block of code.

+ A list of the focal species (.csv file)
+ The PR occurrence records for the hydraulic focal species (.RDA file)
+ The PR SDMs for hydraulic focal species (.grd/.gri files)
+ The PR climate rasters (.grd/.gri files)

```{r download data, eval=F}
download.file("https://github.com/bobmuscarella/PR-spp-PPT-affinity/raw/master/DATA.zip",
              destfile="DATA.zip")
unzip("DATA.zip")
```

The following data is also used but will be downloaded in later parts of the script:

+ Occurrence records from GBIF
+ Chelsa global climate data (.tif files)
+ Occurrence records from BIEN

## Load PR mean annual rainfall raster
```{r env input, eval=F}
ppt <- raster("DATA/PR_mean_annual_ppt.tif")
```


## Load PR occurrence records and extract precipitation values
```{r get occs, eval=F} 
focsp <- read.csv("DATA/Hyraulics_Focal_Species.csv", as.is=T)

focpts <- read.csv("DATA/Hydraulic_species_PR_occurrences.csv")

focpts$PPT <- extract(ppt, focpts[,c('LON','LAT')])

splist <- split(focpts, focpts$scrubbed.name)

pr_df <- data.frame()
for(i in 1:length(splist)){
  r <- raster(paste0("DATA/", focsp$CODE[i], ".grd"))
  opt.ppt <- cellStats((r == cellStats(r, max)) * ppt, max)
  pr_df <- rbind(pr_df, cbind(as.character(focsp$CODE[i]), 
                        names(splist)[i], 
                        median(splist[[i]]$PPT, na.rm=T), 
                        min(splist[[i]]$PPT, na.rm=T), 
                        max(splist[[i]]$PPT, na.rm=T), 
                        opt.ppt,
                        quantile(splist[[i]]$PPT, 0.05, na.rm=T),
                        quantile(splist[[i]]$PPT, 0.95, na.rm=T)))
}

for(j in 3:8) { pr_df[,j] <- round(as.numeric(as.character(pr_df[,j])), 0) }

names(pr_df) <- c("SPCODE", 
                  "SpeciesName",
                  "MedianPPT", 
                  "MinPPT", 
                  "MaxPPT",
                  "OptimalPPT",
                  "5%quantile",
                  "95%quantile")

rownames(pr_df) <- NULL
```

## Get GBIF occurrence data
The `rgbif::occ_search` function can be used to harvest occurrence records from GBIF.  It returns an object of class `gbif` that functions kind of like a list with data returned for each species as an element of the list.  Below, we only take records with coordinates and we set the limit on the number of records per species to 5000 (the default is only 500). This will take a couple of minutes to run...

```{r gbif, eval=F}
occ_gbif <- occ_search(scientificName=focsp$NAME, 
                     limit=5000,
                     hasCoordinate=TRUE)
```

## Clean GBIF coordinates
Many records in GBIF are of questionable quality - especially the location information. There are many ways one can approach 'cleaning' these data but the `CoordinateCleaner` package is pretty slick. Below, we clean the coordinates for each species in turn. The function flags questionable coordinates based on some rules (see `?clean_coordinates` for details) and, in this case, we simply delete any flagged records.  One could investigate each of the flagged records in detail but this is simpler and reproducible.  This also takes a minute to run.

```{r coordinate cleaner, message=F, warning=F, comment=F, results='hide', eval=F}
occ_gbif_cc <- list()

for(i in 1:length(occ_gbif)){
  tmpdata <- data.frame(species=focsp$NAME[i],
                        lon=occ_gbif[[i]]$data$decimalLongitude,
                        lat=occ_gbif[[i]]$data$decimalLatitude)
  
  occ_gbif_cc[[i]] <- clean_coordinates(tmpdata, 
                                        lon="lon",
                                        lat="lat",
                                        value="clean")
}

# Clean up and make them a single data frame
occ_gbif_cc_df <- do.call(rbind, occ_gbif_cc)
```


## Get Chelsa global climate data
We can download the Chelsa global climate data using the `chelsaDL` package, which you need to install from GitHub using the `devtools` package like this: 

`devtools::install_github("matthewkling/chelsaDL")`

Below, we first specify what we want - the climatological means for the bioclimatic variables (all 19 of them).  You can also get past / future predictions, etc.

```{r chelsa data, message=F, eval=F, results='hide'}
# First create the 'order' for what you want to download
ch_data <- ch_queries(variables = c("bio"),
                      layers = 1:19,
                      timeframes = "1979-2013")

# Now download it to a new subfolder
dir.create("DATA/chelsa")
ch_dl(ch_data, dest="DATA/chelsa")

# Read the gridded climate data just downloaded
ch_grids <- stack(paste0("DATA/chelsa/", list.files("DATA/chelsa")))
```

## Extract climate data at GBIF occurrences
```{r gbif chelsa, results='hide', eval=F}
# Extract the Chelsa data at each of the GBIF points
gbif_ch_extract <- extract(ch_grids, occ_gbif_cc_df[,2:3])

# Summarize the data with mean, 95% and 5% quantiles
gbif_ch_means <- apply(gbif_ch_extract, 2, tapply, 
                  occ_gbif_cc_df$species, mean, na.rm=T)
gbif_ch_q95 <- apply(gbif_ch_extract, 2, tapply, 
                occ_gbif_cc_df$species, quantile, 0.95, na.rm=T)
gbif_ch_q05 <- apply(gbif_ch_extract, 2, tapply, 
                occ_gbif_cc_df$species, quantile, 0.05, na.rm=T)

# Rename the columns for convenience
colnames(gbif_ch_means) <- paste0("bio", 1:19, "_mean")
colnames(gbif_ch_q95) <- paste0("bio", 1:19, "_q95")
colnames(gbif_ch_q05) <- paste0("bio", 1:19, "_q05")

# Bind together to a single data.frame
nobs <- as.vector(table(occ_gbif_cc_df$species))
gbif_ch_df <- data.frame(cbind(nobs,
                               gbif_ch_means, 
                               gbif_ch_q95, 
                               gbif_ch_q05))

# Make sure we a row for each focal species
gbif_ch_df <- gbif_ch_df[match(focsp$NAME, 
                               rownames(gbif_ch_df)),]
rownames(gbif_ch_df) <- focsp$NAME
```


## Acquire and process occurrences from BIEN 

We can also harvest data from BIEN. But, for some reason, I get a different number of occurrence records when I use the BIEN r package vs. if I download records from [www.biendata.org](www.biendata.org). In fact, no records are downloaded for CASARB or PREMON! I've asked the package author about this and am waiting for a reply.  In the meantime, the code below works but doesn't return very many records for some species so I would be very hesitant to use it.

```{r, message=F, eval=F}
library(todoBIEN)

# Some names in BIEN are a bit different than we normally use
biennames <- focsp$NAME
biennames[biennames=="Cecropia schreberiana"] <- "Cecropia peltata"
biennames[biennames=="Prestoea montana"] <- "Prestoea acuminata"

occ_bien <- todoBIEN::BIEN_occurrence_species(species=biennames,
                                              user=user,
                                              password=pswd)#,
                                              # observation.type = T, 
                                              # only.geovalid = T, 
                                              # cultivated = T, 
                                              # natives.only = F)

# Remove rows with no coordinates
occ_bien <- occ_bien[!is.na(occ_bien$latitude),]

# Revert species names
occ_bien$scrubbed_species_binomial[occ_bien$scrubbed_species_binomial=="Cecropia peltata"] <- "Cecropia schreberiana"
occ_bien$scrubbed_species_binomial[occ_bien$scrubbed_species_binomial=="Prestoea acuminata"] <- "Prestoea montana"

# Run the coordinate cleaner function
occ_bien_cc <- clean_coordinates(occ_bien, 
                                 species="scrubbed_species_binomial",
                                 lon="longitude",
                                 lat="latitude",
                                 value="clean")

# Extract the climate data at occurrences
bien_ch_extract <- extract(ch_grids, occ_bien_cc[,3:2])

# Summarize the data with mean, 95% and 5% quantiles
bien_ch_means <- apply(bien_ch_extract, 2, tapply, 
                  occ_bien_cc$scrubbed_species_binomial, 
                  mean, na.rm=T)
bien_ch_q95 <- apply(bien_ch_extract, 2, tapply, 
                occ_bien_cc$scrubbed_species_binomial, 
                quantile, 0.95, na.rm=T)
bien_ch_q05 <- apply(bien_ch_extract, 2, tapply, 
                occ_bien_cc$scrubbed_species_binomial, 
                quantile, 0.05, na.rm=T)

# Rename the columns for convenience
colnames(bien_ch_means) <- paste0("bio", 1:19, "_mean")
colnames(bien_ch_q95) <- paste0("bio", 1:19, "_q95")
colnames(bien_ch_q05) <- paste0("bio", 1:19, "_q05")

# Bind together to a single data.frame
nobs <- as.vector(table(occ_bien_cc$scrubbed_species_binomial))
bien_ch_df <- as.data.frame(cbind(nobs,
                                  bien_ch_means, 
                                  bien_ch_q95, 
                                  bien_ch_q05))

# Make sure we a row for each focal species
bien_ch_df <- bien_ch_df[match(focsp$NAME, 
                               rownames(bien_ch_df)),]

rownames(bien_ch_df) <- focsp$NAME

write.csv(bien_ch_df, "BIEN_ppt_affinities.csv")

```
## Combine GBIF and BIEN data before extracting bioclim

```{r combo, eval=F}
names(occ_bien_cc)[c(1,3:2)] <- names(occ_gbif_cc_df)[c(1:3)]

comb_cc_df <- rbind(occ_gbif_cc_df[,c(1:3)], occ_bien_cc[,c(1,3:2)])

# Remove duplicated records?
comb_cc_df <- comb_cc_df[!duplicated(comb_cc_df),]

# Do the extraction
comb_ch_extract <- extract(ch_grids, comb_cc_df[,2:3])

# Summarize the data with mean, 95% and 5% quantiles
comb_ch_median <- apply(comb_ch_extract, 2, tapply, 
                  comb_cc_df$species, 
                  median, na.rm=T)
comb_ch_q95 <- apply(comb_ch_extract, 2, tapply, 
                comb_cc_df$species, 
                quantile, 0.95, na.rm=T)
comb_ch_q05 <- apply(comb_ch_extract, 2, tapply, 
                comb_cc_df$species, 
                quantile, 0.05, na.rm=T)

# Rename the columns for convenience
colnames(comb_ch_median) <- paste0("bio", 1:19, "_median")
colnames(comb_ch_q95) <- paste0("bio", 1:19, "_q95")
colnames(comb_ch_q05) <- paste0("bio", 1:19, "_q05")

# Bind together to a single data.frame
nobs <- as.vector(table(comb_cc_df$species))
comb_ch_df <- as.data.frame(cbind(nobs,
                                  comb_ch_median, 
                                  comb_ch_q95, 
                                  comb_ch_q05))

# Make sure we a row for each focal species
comb_ch_df <- comb_ch_df[match(focsp$NAME, 
                               rownames(comb_ch_df)),]

rownames(comb_ch_df) <- focsp$NAME

write.csv(comb_ch_df, "Combined_BIEN_GBIF_ppt_affinities.csv")

```

## Check precipitation affinity for a few potentially new focal species

```{r}
pts <- readRDS('/Users/au529793/Projects/Postdoc/CLCC/DATA/occurrences/PR_PLANT_OCCS_8.3.16.RDS')

pts$scrubbed.name[pts$scrubbed.name=='Prestoea acuminata'] <- 'Prestoea montana'
pts$scrubbed.name[pts$scrubbed.name=='Myrcia amazonica'] <- 'Myrcia leptoclada'
pts$scrubbed.name[pts$scrubbed.name=='Nectandra turbacensis'] <- 'Ocotea sintenisii'
pts$scrubbed.name[pts$scrubbed.name=='Henriettea squamulosa'] <- 'Calycogonium squamulosum'

pts <- unique(pts)

focsp <- data.frame(sp = c("Calycogonium squamulosum",
                           "Cordia borinquensis",
                           "Cyrilla racemiflora",
                           "Cecropia schreberiana",
                           "Prestoea montana",
                           "Ocotea leucoxylon"),
                    code=c("CALSQU","CORBOR","CYRRAC","CECSCH","PREMON","OCOLEU"))

focpts <- droplevels(pts[pts$scrubbed.name %in% focsp$sp,])

ppt <- raster("DATA/PR_mean_annual_ppt.tif")

focpts$PPT <- extract(ppt, focpts[,c('LON','LAT')])

splist <- split(focpts, focpts$scrubbed.name)

pr_df <- data.frame()
for(i in 1:length(splist)){
  r <- raster(paste0("DATA/", focsp$code[i], ".grd"))
  opt.ppt <- cellStats((r == cellStats(r, max)) * ppt, max)
  pr_df <- rbind(pr_df, cbind(names(splist)[i], 
                        median(splist[[i]]$PPT, na.rm=T), 
                        min(splist[[i]]$PPT, na.rm=T), 
                        max(splist[[i]]$PPT, na.rm=T), 
                        opt.ppt,
                        quantile(splist[[i]]$PPT, 0.05, na.rm=T),
                        quantile(splist[[i]]$PPT, 0.95, na.rm=T)))
}

for(j in 2:7) { pr_df[,j] <- round(as.numeric(as.character(pr_df[,j])), 0) }

names(pr_df) <- c("SpeciesName",
                  "MedianPPT", 
                  "MinPPT", 
                  "MaxPPT",
                  "OptimalPPT",
                  "5%quantile",
                  "95%quantile")

rownames(pr_df) <- NULL

write.csv(pr_df, "/Users/au529793/Desktop/NewTargets_PPTaffinity.csv")

```
